Effectful Semantics for Substructural SSA

### 2023-11-22 pldi24/paper.tex
Work on potential appendix structure
Tap

\begin{proof} 
  We begin by proving \ref{lemma:substitution:term} by induction on derivations \(\hasty{\Gamma}{p}{a}{A}{q}\):
  \begin{itemize}
    \item \rle{var}: we have that \(\cwk{\Gamma}{\thyp{x}{A}{q}}\), and
    therefore that \(\issubst{\gamma}{\Theta}{\thyp{x}{A}{q}}\); by
    \rle{subst-var}, we obtain the desired result that
    \(\hasty{\Theta}{p}{\gamma(x)}{A}{q}\).
    \item \rle{app}: \todo{this}
    \item \rle{pair}: \todo{this}
    \item \rle{true}: \todo{this}
    \item \rle{false}: \todo{this}
    \item \rle{let}: \todo{this}
    \item \rle{unit}: \todo{this}
    \item \rle{let2}: \todo{this}
  \end{itemize}
  Using \ref{lemma:substitution:term}; we can now prove
  \ref{lemma:substitution:block} and \ref{lemma:substitution:cfg} by
  mutual induction on derivations as follows:
  \begin{itemize}
    \item \rle{br}: \todo{this}
    \item \rle{let-blk}: \todo{this}
    \item \rle{ite}: \todo{this}
    \item \rle{let2-blk}: \todo{this}
    \item \rle{where}: \todo{this}
    \item \rle{nil-br}: \todo{this}
    \item \rle{cons-br}: \todo{this}
  \end{itemize}
\end{proof}

@@

% \subsection{State-Passing Category}
%
% We can further use this example to get a substructural \textit{monoidal} category by considering a primitive form of separation logic: let's introduce a grammar of predicates \(\varphi, \psi ::= \top, \ms{S}, \bot\) for programs which cannot access state, can access state, and are invalid, respectively. The separating conjunction operator on these predicates is then given by
% \begin{equation}
%   \top * \varphi = \varphi * \top = \varphi
%   \qquad
%   \bot * \varphi = \varphi * \bot = \bot
%   \qquad
%   \ms{S} * \ms{S} = \bot
% \end{equation}
% We'll then define a category \(\ms{SRel}\) with objects of the form \((A, \varphi)\) and morphisms
% \begin{equation}
%   \begin{gathered}
%     \boxed{
%       \ms{SRel}((A, \varphi), (B, \psi)) 
%       \subseteq A \to \ms{S} \to \mc{P}(A \times S) 
%     }
%     \\
%     \begin{aligned}
%       \ms{SRel}((A, \top), (B, \top)) &= \{f \in A \to S \to \mc{P}(A \times S) \mid \exists g. f = \lambda a, s. g(a) \times \{s\}\} \\
%       \ms{SRel}((A, \ms{S}), (B, \ms{S})) &= A \to S \to \mc{P}(A \times S) \\
%       \ms{SRel}((A, \varphi), (B, \psi)) &= \{\lambda a, s. \varnothing\} \quad \text{otherwise}
%     \end{aligned}
%   \end{gathered}
% \end{equation}
% Note that it is precisely the \textit{central} morphisms \(\ms{C}_0(A, B)\) which lie in \(\ms{SRel}((A, \top), (B, \top))\). We can then define the tensor product on objects as \((A, \varphi) \otimes (B, \psi) = (A \otimes B, \varphi * \psi)\), and, for \(f \in \mathsf{SRel}((A, \varphi), (B, \psi))\), and tensor product functors
% \begin{equation}
%   \begin{aligned}
%   f \otimes (C, \phi) &= f \otimes C,
%   & (C, \phi) \otimes f &= C \otimes f
%   && \text{if}\; \varphi * \phi \neq \bot, \psi * \phi \neq \bot \\
%   f \otimes (C, \phi) &= \lambda a, s. \varnothing,
%   & (C, \phi) \otimes f &=  \lambda a, s. \varnothing
%   && \text{otherwise}
%   \end{aligned} 
% \end{equation}
% Lifting associators, unitors, and symmetries from \(\mc{C}_0\) in the obvious manner (setting them to \(\lambda a, s. \varnothing\) where necessary), we can verify that \(\ms{SRel}\) is a \textit{premonoidal} category. we can see that it is in fact \textit{monoidal} by nothing that, for any \(f, g\), \(f \ltimes g = f \rtimes g\) since either \(f\) is central, \(g\) is central, or their product is the empty morphism \(\lambda a, s. \varnothing\). Hence, to get a \textit{substructural} monoidal category, we simply need to define:
% \begin{equation}
%   \begin{gathered}
%   \ms{Aff}(\ms{SRel}) = \ms{Rel}(\ms{SRel}) = \{(A, \top) | A \in \ms{Set}\} \neq |\ms{SRel}|
%   \\
%   \begin{aligned}
%     \ms{SRel}^\varnothing((A, \phi), (B, \psi))
%     &= \ms{SRel}((A, \phi), (B, \psi)) \\
%     \ms{SRel}^{\{\ms{a}\}}((A, \phi), (B, \psi)) 
%     &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| \geq 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
%     \ms{SRel}^{\{\ms{r}\}}((A, \phi), (B, \psi)) 
%     &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| \leq 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
%     \implies \ms{SRel}^{\{\ms{a}, \ms{r}\}}((A, \phi), (B, \psi)) 
%     &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| = 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
%   \end{aligned}
%   \end{gathered}
% \end{equation}
% We can verify that the associators, unitors, and symmetry lie in \(\ms{SRel}^{\{\ms{a}, \ms{r}\}}\) and that all other desired axioms are satisfied.

>>>>

\begin{proof} 
  We begin by proving \ref{lemma:substitution:term} by induction on derivations \(\hasty{\Gamma}{p}{a}{A}{q}\):
  \begin{itemize}
    \item \rle{var}: we have that \(\cwk{\Gamma}{\thyp{x}{A}{q}}\), and
    therefore that \(\issubst{\gamma}{\Theta}{\thyp{x}{A}{q}}\); by
    \rle{subst-cons}, we obtain the desired result that
    \(\hasty{\Theta}{p}{\gamma(x)}{A}{q}\).
    \item \rle{app}: by induction, we have that
    \(\hasty{\Theta}{1}{[\gamma]a}{A}{q}\), and therefore we can apply \rle{app}
    to obtain \(\hasty{\Theta}{p}{[\gamma](f\;a)}{B}{q}\) as desired.
    \item \rle{pair}: we have that \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\) and
    \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). Therefore, by induction, we have that
    \(\hasty{\Theta_\Delta}{1}{[\gamma]a}{A}{q}\) and
    \(\hasty{\Theta_\Xi}{1}{[\gamma]b}{B}{q}\); we can simply apply \rle{pair}
    to obtain our desired result that \(\hasty{\Theta}{p}{[\gamma](a, b)}{A
    \otimes B}{q}\)
    \item \rle{true}, \rle{unit}, \rle{false}: these cases hold trivially since
    \(\cwk{\Theta}{\cdot}\)
    \item \rle{let}: we have that \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
    and \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). Taking \(x\) to be a fresh
    variable (with \(\gamma(x) = x\)), we can conclude by \rle{subst-cons} that
    \(\issubst{\gamma}{\Theta_\Delta, \thyp{x}{A}{}}{\Delta, \thyp{x}{A}{}}\).
    Therefore, by induction, we have that \(\hasty{\Theta_\Delta, \thyp{x}{A}{}}{p}{[\gamma]e}{B}{q}\) and \(\hasty{\Theta_\Xi}{1}{[\gamma]a}{A}{q}\); we
    can hence simply apply \rle{let} to obtain
    \(\hasty{\Theta}{p}{[\gamma]\letexpr{x}{a}{e}}{B}{q}\), as desired.
    \item \rle{let2}: we have that \(\issubst{\gamma}{\Theta_\Delta}{\Delta}\)
    and \(\issubst{\gamma}{\Theta_\Xi}{\Xi}\). Taking \(x, y\) to be fresh
    variables (with \(\gamma(x) = x\), \(\gamma(y) = y\)), we can conclude by
    \rle{subst-cons} that \(\issubst{\gamma}{\Theta_\Delta, \thyp{x}{A}{},
    \thyp{y}{B}{}}{\Delta, \thyp{x}{A}{}, \thyp{y}{B}{}}\). Therefore, by
    induction, we have that \(\hasty{\Theta_\Delta, \thyp{x}{A}{},
    \thyp{y}{B}{}}{p}{[\gamma]e}{C}{q}\) and
    \(\hasty{\Theta_\Xi}{1}{[\gamma]a}{A \otimes B}{q}\); we can hence simply
    apply \rle{let2} to obtain
    \(\hasty{\Theta}{p}{[\gamma]\letexpr{(x, y)}{a}{e}}{C}{q}\), as desired.
  \end{itemize}
  Using \ref{lemma:substitution:term}; we can now prove
  \ref{lemma:substitution:block} and \ref{lemma:substitution:cfg} by mutual
  induction on derivations as follows:
  \begin{itemize}
    \item \rle{br}: \todo{this}
    \item \rle{let-blk}: \todo{this}
    \item \rle{ite}: \todo{this}
    \item \rle{let2-blk}: \todo{this}
    \item \rle{where}: \todo{this}
    \item \rle{nil-br}: \todo{this}
    \item \rle{cons-br}: \todo{this}
  \end{itemize}
\end{proof}

@@

\subsection{Equational Theory}

\TODO{\(\beta\), \(\eta\), etc}

\section{Refinement}

In general, we often want to justify optimizations which \textit{refine} a
program by, for example, reducing the set of allowable behaviours, rather than
simply rewrite the program to an equivalent one.

\TODO{freeze example, UB}

\TODO{determinism and local refinement}

\subsection{Poset-Enriched Categories}

\TODO{semantics of refinement via poset-enriched categories}

\TODO{trivial poset-enrichment}

\TODO{powerset-based poset-enrichment}

\subsection{Semantic Refinement}

\TODO{refinement for rewrites}

\TODO{semantic refinement theorem}

\subsection{Equational Theory}

\TODO{this}

\section{Normalisation}

\TODO{this}

\subsection{Strict SSA form}

\TODO{this}

\subsection{Structured Control-Flow}

\TODO{this}

\section{Loop Optimizations}

\TODO{this}

\subsection{Unrolling and Vectorization}

\TODO{this}

\subsection{Dead-code Elimination}

\TODO{this}

\subsection{Loop Hoisting}

\TODO{this}

\subsection{Loop Fusion/Fission}

\TODO{this}

\section{E-Graph Optimizations}

\TODO{this}

\subsection{Floating SSA}

\TODO{this}

\subsection{Whole-program rewriting}

\TODO{this}

% \section{State-Passing Category}
%
% We can further use this example to get a substructural \textit{monoidal} category by considering a primitive form of separation logic: let's introduce a grammar of predicates \(\varphi, \psi ::= \top, \ms{S}, \bot\) for programs which cannot access state, can access state, and are invalid, respectively. The separating conjunction operator on these predicates is then given by
% \begin{equation}
%   \top * \varphi = \varphi * \top = \varphi
%   \qquad
%   \bot * \varphi = \varphi * \bot = \bot
%   \qquad
%   \ms{S} * \ms{S} = \bot
% \end{equation}
% We'll then define a category \(\ms{SRel}\) with objects of the form \((A, \varphi)\) and morphisms
% \begin{equation}
%   \begin{gathered}
%     \boxed{
%       \ms{SRel}((A, \varphi), (B, \psi)) 
%       \subseteq A \to \ms{S} \to \mc{P}(A \times S) 
%     }
%     \\
%     \begin{aligned}
%       \ms{SRel}((A, \top), (B, \top)) &= \{f \in A \to S \to \mc{P}(A \times S) \mid \exists g. f = \lambda a, s. g(a) \times \{s\}\} \\
%       \ms{SRel}((A, \ms{S}), (B, \ms{S})) &= A \to S \to \mc{P}(A \times S) \\
%       \ms{SRel}((A, \varphi), (B, \psi)) &= \{\lambda a, s. \varnothing\} \quad \text{otherwise}
%     \end{aligned}
%   \end{gathered}
% \end{equation}
% Note that it is precisely the \textit{central} morphisms \(\ms{C}_0(A, B)\) which lie in \(\ms{SRel}((A, \top), (B, \top))\). We can then define the tensor product on objects as \((A, \varphi) \otimes (B, \psi) = (A \otimes B, \varphi * \psi)\), and, for \(f \in \mathsf{SRel}((A, \varphi), (B, \psi))\), and tensor product functors
% \begin{equation}
%   \begin{aligned}
%   f \otimes (C, \phi) &= f \otimes C,
%   & (C, \phi) \otimes f &= C \otimes f
%   && \text{if}\; \varphi * \phi \neq \bot, \psi * \phi \neq \bot \\
%   f \otimes (C, \phi) &= \lambda a, s. \varnothing,
%   & (C, \phi) \otimes f &=  \lambda a, s. \varnothing
%   && \text{otherwise}
%   \end{aligned} 
% \end{equation}
% Lifting associators, unitors, and symmetries from \(\mc{C}_0\) in the obvious manner (setting them to \(\lambda a, s. \varnothing\) where necessary), we can verify that \(\ms{SRel}\) is a \textit{premonoidal} category. we can see that it is in fact \textit{monoidal} by nothing that, for any \(f, g\), \(f \ltimes g = f \rtimes g\) since either \(f\) is central, \(g\) is central, or their product is the empty morphism \(\lambda a, s. \varnothing\). Hence, to get a \textit{substructural} monoidal category, we simply need to define:
% \begin{equation}
%   \begin{gathered}
%   \ms{Aff}(\ms{SRel}) = \ms{Rel}(\ms{SRel}) = \{(A, \top) | A \in \ms{Set}\} \neq |\ms{SRel}|
%   \\
%   \begin{aligned}
%     \ms{SRel}^\varnothing((A, \phi), (B, \psi))
%     &= \ms{SRel}((A, \phi), (B, \psi)) \\
%     \ms{SRel}^{\{\ms{a}\}}((A, \phi), (B, \psi)) 
%     &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| \geq 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
%     \ms{SRel}^{\{\ms{r}\}}((A, \phi), (B, \psi)) 
%     &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| \leq 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
%     \implies \ms{SRel}^{\{\ms{a}, \ms{r}\}}((A, \phi), (B, \psi)) 
%     &= \{f \in \ms{SRel}((A, \phi), (B, \psi)) \mid \forall a, s. |f(a, s)| = 1 \lor \phi \neq \psi \lor \phi = \bot \} \\
%   \end{aligned}
%   \end{gathered}
% \end{equation}
% We can verify that the associators, unitors, and symmetry lie in \(\ms{SRel}^{\{\ms{a}, \ms{r}\}}\) and that all other desired axioms are satisfied.
